Equalizer 2.0 --\\Convergence of a Parallel Rendering Framework

### 2018-02-08 paperIEEE.tex
Fix tex
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@109 db7c848b-7755-4df2-b763-8796797afa52
Collage review
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@110 db7c848b-7755-4df2-b763-8796797afa52

The region of interest is the screen-space 2D bounding box enclosing the data
rendered by a single resource. We have extended the core parallel rendering
framework to use an application-provided ROI to optimize the
\texsf{load\_equalizer} as well as the image compositing. The load equalizer
uses the ROI to refine its load grid to the regions containing data. The
compositing code uses the ROI to minimize image readback and network
transmission. In \cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of
the algorithm, and show that using ROI can quadruple the rendering performance,
in particular for the costly compositing step in sort-last rendering.

@@

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved screens.
These frusta describe a physically-correct display setup for a Virtual Reality
installation. \label{sec:swap} A canvas may have a software or hardware swap
barrier, which will synchronize the rendering of all contributing GPUs. The
software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all
segments, the swap buffers call followed by a \textsf{glFlush} to ensure timely
execution of the swap command. The hardware swap barrier is implemented using a
\textsf{NV\_swap\_group} extension.

@@

\fig{fLayout} shows RTT Deltagen running an example multi-segment, multi-view
setup driven by eight rendering nodes. The main tiled display wall canvas uses
four LCD segments showing one layout with four views, which do not align on the
segment boundaries. This setup creates seven destination channels. The
configuration provides multiple, run-time configurable layouts. It is driven
from the control host on the right, which shows four views, each in their own
canvas and segment windows with a single-view layout each. One view on the
control host synchronizes its content (model and camera) with one view on the
display wall through Collage objects. The control host allows full model
modifications and all workflows supported within the standalone Deltagen
application, and all changes are synchronized to the corresponding rendering
nodes. Since this setup is monoscopic, no head tracking or observers are used.

@@

This versatile mechanism allows non-experts to configure and profit from
multi-GPU workstations and visualization clusters, as well as to provide system
administrators with the tools to implement easy to use integration with cluster
schedulers. This feature is transparent to application developers.

@@

Qt is a popular window system with many application developers. Unfortunately,
it imposes a different threading model for window creation and event handling
compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread
is independently responsible for creating its windows, receiving the events and
eventually dispatching them to the application process main thread. This design
is motivated by the natural threading model of X11 and WGL, and allows simple
sequential semantics between OpenGL rendering and event handling. In contrast,
Qt requires all windows and QOpenGLContext to be created from the Qt main
thread. An existing Qt window or context may subsequently be moved to a
different thread, and events are signalled from the main thread. For Qt windows,
Equalizer will internally dispatch and handle the window creation from the
render to the main thread, move the created objects back to the render thread,
and dispatch Qt signals to the correct render threads.

@@

\begin{compactdesc}
\item[Connection] A stream-oriented point-to-point communication
  line. The connections
  transmit raw data reliably between two endpoints for unicast connections, and
  between a set of endpoints for multicast connections. For unicast,
  process-local pipes, TCP and Infiniband RDMA are implemented. For multicast,
  a reliable, UDP-based protocol is discussed in \sref{sec:RSP}.
\item[DataI/OStream] Abstracts the input and output of C++ data types from or to
  a set of connections by implementing output stream operators. Uses buffering
  to aggregate data for network transmission. Performs endian swapping during
  input if the endianness differs between the originating and local node.
\item[Node and LocalNode] The abstraction of a process in the cluster. Nodes
  communicate with each other using connections. A LocalNode listens on various
  connections and processes requests for a given process. Received data is
  wrapped in ICommands and dispatched to command handler methods. A Node is a
  proxy for a remote LocalNode. The \textsf{Equalizer} Client object is a
LocalNode.
\item[Object] Provides object-oriented, versioned data distribution of C++
  objects between nodes. Objects are registered or mapped on a Local\-Node.
\end{compactdesc}

@@

RSP is an implementation of a reliable multicast protocol over unreliable UDP
transport. RSP behaves similarly to TCP, it provides full reliability and
ordering of the data, and slow receivers will eventually throttle the sender
through a sliding window algorithm. This behavior is needed to guarantee
delivery of data in all situations. Pragmatic generic multicast (PGM~\cite{pgm})
provides full ordering, but slow clients will disconnect from the multicast
session instead of throttling the send rate.

@@

Our RSP implementation uses a separate protocol thread for each RSP group, which
handles all reads and writes on the multicast group. This thread implements the
protocol handling and communicates with the application threads through
thread-safe queues. These queues contain datagrams of the application byte
stream, prefixed by a type-specific header of at most eight bytes. Each
connection has a configurable number of buffers, (1024 by default) of a
configurable MTU (1470 bytes default) which are either free for use or in
transmission.

Handling a smooth packet flow is critical for performance. RSP uses active flow
control to advance the byte stream buffered by the implementation. Each incoming
connection actively acknowledges every $n$ (17 by default) packets fully
received. The incoming connections temporarily offset this acknowledgment by
their connection identifier to avoid bursts of acks. Any missed datagram is
actively nack'ed as soon as it is detected. Write connections continuously
retransmit packets from nack datagrams, and advance their window upon reception
of acks from all readers. Furthermore, the writer will explicitly request an ack
or nack when it runs out of empty buffers or finishes its write queue. Nack
datagrams may contain multiple ranges of missed datagrams, which is motivated by
the observation that UDP implementations often drop multiple packets at once.

Congestion control is necessary to optimize bandwidth usage. While TCP uses the
well-known AIMD (additive increase, multiplicative decrease) algorithm, we have
chosen a more aggressive congestion control algorithm of additive increase and
additive decrease. This has proven experimentally to be more optimal: UDP is
often rate-limited by switches; that is, packets are discarded regularly and not
exceptionally. Only slowly backing of the current send rate helps to stay close
to this limit. Furthermore, our RSP traffic is limited to the local subnet,
making cooparation between multiple data stream less of an issue. Send rate
limiting uses a bucket algorithm, where over time the bucket fills with send
credits, from which sends are substracted. If there are no available credits,
the sender sleeps until sufficient credits are available.

@@

Adapting an existing application for parallel rendering requires the
synchronization of application data across the processes in the parallel
rendering setup. Existing parallel rendering frameworks address this often
poorly, at best they rely on MPI to distribute data. Real-world, interactive
visualization applications are typically written in C++ and have complex data
models and class hierarchies to represent their application state. As outlined
in \cite{EMP:09}, the parallel rendering code in an \textsf{Equalizer}
application only
needs access to the data needed for rendering, as all application logic is
centralized in the application main thread. We have encountered two main
approaches to address this distribution: Using a shared filesystem for static
data, or using data distribution for static and dynamic data.

Distributed objects are not required to build \textsf{Equalizer} applications.
While most
developers choose to use this abstraction for convenience, we have seen
applications using other means for data distribution, e.g., MPI.

@@

Distributed objects in \textsf{Collage} provide powerful, object-oriented data
distribution for C++ objects. They facilitate the implementation of data
distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
application programmer implements serialization and deserialization of the
distributed data. Distributed objects can be static (immutable) or
dynamic. Objects have a universally unique identifier (UUID) to address them
cluster-wide. A master-slave model is used to establish mapping and data
synchronization across processes. Typically, the application main loop registers
a master instance and communicates the UUID to the render clients, which map
their instance to the given identifier. The following object types are
available:

@@

All distributed objects have to implement \textsf{getInstanceData} and
\textsf{applyInstanceData} to serialize and deserialize the object's distributed
data. These methods provide an output or input stream as a parameter, which
abstracts the data transmission and can be used like a \textsf{std::stream}.
The data streams implement efficient buffering and compression, and
automatically select the best connection for data transport. Custom data type
serializers can be implemented by providing the appropriate serialization
functions. No pointers should be directly transmitted through the data
streams. For pointers, the corresponding object is typically a distributed
object as well, and its identifier and potentially version is transmitted in
place of the memory address.

Dynamic objects are versioned, and have to override \textsf{getChangeType} to
indicate how they want to have changes to be handled. Upon \textsf{commit} the
delta data from the previous version is sent, if available using multicast, to
all mapped slave instances. The data is queued on the remote node, and is
applied when the application calls \textsf{sync} to synchronize the object to a
new version. The \textsf{sync} method might block if a version has not yet been
committed or is still in transmission. All versioned objects the following
characteristics:

@@

Besides the instance data serialization methods used to map an object, delta and
unbuffered objects may implement \textsf{pack} and \textsf{unpack} to serialize
or deserialize the changes since the last version.

@@

Blocking commits allow to limit the number of outstanding, queued versions on
the slave nodes. A token-based protocol will block the commit on the master
instance if to many unsynchronized versions exist.

@@

The API presented in the previous section provides sufficient abstraction to
implement various optimizations for faster mapping and synchronization of data:
compression, chunking, caching, preloading and multicast. The results section
evaluates these optimizations.

@@

The data streaming interface also implements chunking, which aims to pipeline
the serialization code with the network transmission. After a configurable
number of bytes has been serialized to the internal buffer, it is transmitted
and serialization continues. This is used both for the initial mapping data and
for commit data.

Caching retains instance data of objects in a cache, and reuses this data to
accelarate mapping of objects. The instance cache is either filled by
``snooping'' on multicast transmissions or by an explicit preloading when the
master objects are registered. The preloading (``send on register'') sends
instance data of recently registered master objects while the correspinding node
is idle to all connected nodes. These nodes simply enter the received data to
their cache. Preloading uses multicast when available.

@@

During data synchronization, the master instance commits the object data to all
mapped slave instances simultaneously. This is a ``push'' operation, whereas the
mapping is a slave-triggered ``pull'' operation with additional synchronization
points. Slave nodes queue this data and consume it during synchronization. We
test the time to commit and sync the data using different compression engines.

@@

The volume data sets are distributed in a single object, serializing the raw
volume buffer. The Spike volume data set has a significant compression ratio,
which is reflected by the results in \fig{fSpikesDist}. Compression for this
data is beneficial for transmitting data over a 10~Gb/s link even for a single
slave process only. Buffering has little benefit since the serialization of
volume data is trivial. Buffered compression makes a significant difference,
since the compression cost can be amortized over $n$ nodes, reaching raw data
transmission rates of 3.7 GB/s with the default Snappy compressor and at best
4.4 GB/s with ZStandard at level 1.

@@

Unsurprisingly, TCP over the faster InfiniBand link outperforms the cheaper ten
gigabit ethernet link by more than a factor of two. Unexpectedly, the native
RDMA connection performs worse, even though it outperforms IPoIB in a simple
peer-to-peer connection benchmark. This needs further investigation, but we
suspect the abstraction of a byte stream connection chosen by \textsf{Collage}
for historical reasons is not well suited for remote DMA semantics; that is,
one needs to design the network API around zero-copy semantics with managed
memory for modern high-speed transports. Both Infiniband connections show
significant measurement jitter.

Our RSP multicast implementation performs as expected. \textsf{Collage} starts
using multicast to commit new object versions when two or more clients are
mapped, since the transmission to a single client is better done using unicast.
RSP consistently outperforms unicast on the same physical interface and shows
good scaling behavior (2.5 times slower on 16 vs 2 clients on ethernet, 1.8
times slower on InfiniBand). The scaling is significantly better when only one
process per node is used (\fig{fIFDist}, middle: 30\% slower on ethernet,
nearly flat on InfiniBand). The increased transmission time with multiple
clients is caused by a higher probability of packet loss, which increases
significantly when using more than one process per node, as shown in
\fig{fIFDist}, right. This figure plots the number of retransmissions
divided by the number of datagrams. Infiniband outperforms ethernet slightly,
but is largely limited by the RSP implementation throughput of preparing and
queueing the datagrams to and from the protocol thread, which we observed in a
profiling analysis.

@@

Furthermore, we present many new features needed in parallel rendering
applications, from advanced Virtual Reality support to advanced display system
setup for 2D/3D integration, auto-configuration and runtime reconfiguration, and
an advanced network data synchronization library tailored to parallel rendering
applications. We highlight a few commercial and research applications
underlining the generic and versatile system implemenation.

>>>>

The region of interest is the screen-space 2D bounding box enclosing the data
rendered by a single resource. We have extended the core parallel rendering
framework to use an application-provided ROI to optimize the
\textsf{load\_equalizer} as well as the image compositing. The load equalizer
uses the ROI to refine its load grid to the regions containing data. The
compositing code uses the ROI to minimize image readback and network
transmission. In \cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of
the algorithm, and show that using ROI can quadruple the rendering performance,
in particular for the costly compositing step in sort-last rendering.

@@

A canvas can define a frustum, which will create default, planar sub-frusta for
all of its segments. A segment can also define a frustum, which overrides the
canvas frustum, e.g., for non-planar setups such as CAVEs or curved screens.
These frusta describe a physically-correct display setup for a Virtual Reality
installation. \label{sec:swap} A canvas may have a software or hardware swap
barrier, which will synchronize the rendering of all contributing GPUs. The
software barrier executes a \textsf{glFinish} to ensure the GPU is ready to
swap, a \textsf{Collage} barrier (\sref{sec:barrier}) to synchronize all
segments, the swap buffers call followed by a \textsf{glFlush} to ensure timely
execution of the swap command. The hardware swap barrier is implemented using
the \textsf{NV\_swap\_group} extension.

@@

\fig{fLayout} shows RTT Deltagen running an example multi-segment, multi-view
setup driven by eight rendering nodes. The main tiled display wall canvas uses
four LCD segments showing one layout with four views, which do not align on the
segment boundaries. This setup creates seven destination channels. The
configuration provides multiple, run-time configurable layouts. It is driven
from the control host on the right, which shows four views, each in their own
canvas and segment windows with a single-view layout each. One view on the
control host synchronizes its content (model and camera) with one view on the
display wall through Collage objects. The control host allows full model
modifications and all workflows supported within the standalone Deltagen
application, and all changes are synchronized to the corresponding rendering
nodes. For this monoscopic setup no head tracking or observers are used.

@@

This versatile mechanism allows non-experts to configure and profit from
multi-GPU workstations and visualization clusters, as well as to provide system
administrators with the tools to implement easy to use integration with cluster
schedulers.

@@

Qt is a popular window system with application developers. Unfortunately, it
imposes a different threading model for window creation and event handling
compared to \textsf{Equalizer}. In \textsf{Equalizer}, each GPU rendering thread
is independently responsible for creating its windows, receiving the events and
eventually dispatching them to the application process main thread. This design
is motivated by the natural threading model of X11 and WGL, and allows simple
sequential semantics between OpenGL rendering and event handling. In contrast,
Qt requires all windows and QOpenGLContext to be created from the Qt main
thread. An existing Qt window or context may subsequently be moved to a
different thread, and events are signalled from the main thread. For Qt windows,
Equalizer will internally dispatch and handle the window creation from the
render to the main thread, move the created objects back to the render thread,
and dispatch Qt signals to the correct render threads.

@@

\begin{compactdesc}
\item[Connection] A stream-oriented point-to-point communication
  line. The connections
  transmit raw data reliably between two endpoints for unicast connections, and
  between a set of endpoints for multicast connections. For unicast,
  process-local pipes, TCP and Infiniband RDMA are implemented. For multicast,
  a reliable, UDP-based protocol is discussed in \sref{sec:RSP}.
\item[DataI/OStream] Abstracts the input and output of C++ data types from or to
  a set of connections by implementing output stream operators. Uses buffering
  to aggregate data for network transmission. Performs byte swapping during
  input if the endianness differs between the remote and local node.
\item[Node and LocalNode] The abstraction of a process in the cluster. Nodes
  communicate with each other using connections. A LocalNode listens on various
  connections and processes requests for a given process. Received data is
  wrapped in ICommands and dispatched to command handler methods. A Node is a
  proxy for a remote LocalNode. The \textsf{Equalizer} Client object is a
LocalNode.
\item[Object] Provides object-oriented, versioned data distribution of C++
  objects between nodes. Objects are registered or mapped on a Local\-Node.
\end{compactdesc}

@@

RSP is an implementation of a reliable multicast protocol over unreliable UDP
transport. RSP behaves similarly to TCP; it provides full reliability and
ordering of the data, and slow receivers will eventually throttle the sender
through a sliding window algorithm. This behavior is needed to guarantee
delivery of data in all situations. Pragmatic generic multicast (PGM~\cite{pgm})
provides full ordering, but slow clients will disconnect from the multicast
session instead of throttling the send rate.

@@

Our RSP implementation uses a separate protocol thread for each RSP group, which
handles all reads and writes on the multicast socket. It implements the protocol
handling and communicates with the application threads through thread-safe
queues. The queues contain datagrams of the application byte stream, prefixed by
a header of at most eight bytes. Each connection has a configurable number of
buffers (1024 by default) of a configurable MTU (1470 bytes default), which are
either free or in transmission.

Handling a smooth packet flow is critical for performance. RSP uses active flow
control to advance the byte stream buffered by the implementation. Each incoming
connection actively acknowledges every $n$ (17 by default) packets fully
received. The incoming connections offset this acknowledgment by their
connection identifier to avoid bursts of acks. Any missed datagram is actively
nack'ed as soon as detected. Write connections continuously retransmit packets
for nack datagrams, and advance their window upon reception of all acks from the
group. The writer will explicitly request an ack or nack when it runs out of
empty buffers or finishes its write queue. Nack datagrams may contain multiple
ranges of missed datagrams, which is motivated by the observation that UDP
implementations often drop multiple contiguous packets.

Congestion control is necessary to optimize bandwidth usage. While TCP uses the
well-known additive increase, multiplicative decrease algorithm, we have chosen
a more aggressive congestion control algorithm of additive increase and additive
decrease. This has proven experimentally to be more optimal: UDP is often
rate-limited by switches; that is, packets are discarded regularly and not
exceptionally. Only slowly backing of the current send rate helps to stay close
to this limit. Furthermore, our RSP traffic is limited to the local subnet,
making cooperation between multiple data stream less of an issue. Send rate
limiting uses a bucket algorithm, where over time the bucket fills with send
credits, from which sends are substracted. If there are no available credits,
the sender sleeps until sufficient credits are available.

@@

Adapting an existing application for parallel rendering requires the
synchronization of application data across the processes in the parallel
rendering setup. Existing parallel rendering frameworks address this often
poorly, at best they rely on MPI to distribute data. Real-world, interactive
visualization applications are typically written in C++ and have complex data
models and class hierarchies to represent their application state. As outlined
in \cite{EMP:09}, the parallel rendering code in an \textsf{Equalizer}
application only needs access to the data needed for rendering, as all
application logic is centralized in the application main thread. We have
encountered two main approaches to address this distribution: Using a shared
filesystem for static data, or using data distribution for static and dynamic
data. Distributed objects are not required to build \textsf{Equalizer}
applications. While most developers choose to use this abstraction for
convenience, we have seen applications using other means for data distribution,
e.g., MPI.

@@

Distributed objects in \textsf{Collage} provide powerful, object-oriented data
distribution for C++ objects. They facilitate the implementation of data
distribution in a cluster environment. Distributed objects are created by
subclassing from \textsf{co::Serializable} or \textsf{co::Object}. The
application programmer implements serialization and deserialization. Distributed
objects can be static (immutable) or dynamic. Objects have an universally unique
identifier (UUID) as cluster-wide addresses. A master-slave model is used to
establish mapping and data synchronization across processes. Typically, the
application main loop registers a master instance and communicates the UUID to
the render clients, which map their instance to the given identifier. The
following object types are available:

@@

Serialization is facilitated using output or input streams, which abstract the
data transmission and are used like a \textsf{std::stream}. The data streams
implement efficient buffering and compression, and automatically select the best
connection for data transport. Custom data type serializers can be implemented
by providing the appropriate serialization functions. No pointers should be
directly transmitted through the data streams. For pointers, the corresponding
object is typically a distributed object as well, and its UUID and version are
transmitted in place of a pointer.

Dynamic objects are versioned, and on \textsf{commit} the delta data from the
previous version is sent, if available using multicast, to all mapped slave
instances. The data is queued on the remote node, and is applied when the
application calls \textsf{sync} to synchronize the object to a new version. The
\textsf{sync} method might block if a version has not yet been committed or is
still in transmission. All versioned objects the following characteristics:

@@

@@

Blocking commits allow to limit the number of outstanding, queued versions on
the slave nodes. A token-based protocol will block the commit on the master
instance if too many unsynchronized versions exist.

@@

The API presented in the previous section provides sufficient abstraction to
implement various optimizations for faster mapping and synchronization of data:
compression, chunking, caching, preloading and multicast. The results section
evaluates some of these optimizations.

@@

The data streaming interface implements chunking, which pipelines the
serialization code with the network transmission. After a configurable number of
bytes has been serialized to the internal buffer, it is transmitted and
serialization continues. This is used both for the initial mapping data and for
commit data.

Caching retains instance data of objects in a client-side cache, and reuses this
data to accelerate mapping of objects. The instance cache is either filled by
``snooping'' on multicast transmissions or by an explicit preloading when the
master objects are registered. The preloading sends instance data of recently
registered master objects while the corresponding node is idle to all connected
nodes. These nodes simply enter the received data to their cache. Preloading
uses multicast when available.

@@

During data synchronization, the master commits the object data to all mapped
slave instances simultaneously. This is a ``push'' operation, whereas the
mapping is a slave-triggered ``pull'' operation. Slave nodes queue this data and
consume it during synchronization. We test the time to commit and sync the data
using different compression engines.

@@

The volume data sets are distributed in a single object, serializing the raw
volume buffer. The Spike volume data set has a significant compression ratio,
which is reflected by the results in \fig{fSpikesDist}. Compression for this
data is beneficial for transmitting data over a 10~Gb/s link even for a single
slave process. Buffering has little benefit since the serialization of volume
data is trivial. Buffered compression makes a significant difference, since the
compression cost can be amortized over $n$ nodes, reaching raw data transmission
rates of 3.7 GB/s with the default Snappy compressor and at best 4.4 GB/s with
ZStandard at level 1.

@@

Unsurprisingly, TCP over the faster InfiniBand link outperforms the cheaper ten
gigabit ethernet link by more than a factor of two. Unexpectedly, the native
RDMA connection performs worse, even though it outperforms IPoIB in a simple
peer-to-peer connection benchmark. This needs further investigation, but we
suspect the abstraction of a byte stream connection chosen by \textsf{Collage}
is not well suited for remote DMA semantics; that is, one needs to design the
network API around zero-copy semantics with managed memory for modern high-speed
transports. Both Infiniband connections show significant measurement jitter.

RSP multicast performs as expected. \textsf{Collage} starts using multicast to
commit new object versions when two or more clients are mapped, since the
transmission to a single client is faster using unicast. RSP consistently
outperforms unicast on the same physical interface and shows good scaling
behavior (2.5 times slower on 16 vs 2 clients on ethernet, 1.8 times slower on
InfiniBand). The scaling is significantly better when only one process per node
is used (\fig{fIFDist}, middle: 30\% slower on ethernet, nearly flat on
InfiniBand). The increased transmission time with multiple clients is caused by
a higher probability of packet loss, which increases significantly when using
more than one process per node. \fig{fIFDist} (right) plots the number of
retransmissions divided by the number of datagrams. Infiniband outperforms
ethernet slightly, but is largely limited by the RSP implementation throughput
of preparing and queueing the datagrams to and from the protocol thread, which
we observed in profiling.

@@

Furthermore, we present many new features needed in parallel rendering
applications, from advanced Virtual Reality support to advanced display system
setup for 2D/3D integration, auto-configuration and runtime reconfiguration, and
an advanced network data synchronization library tailored to parallel rendering
applications. We highlight a few commercial and research applications
underlining the generic and versatile system implementation.
